---
title: "Conditional Exchangability"
author: "Uriah Finkel"
format: revealjs
---

## Causal Question {.smaller}

```{r}
#| echo: false

library(reactable)
library(dplyr)
```

::: columns
::: {.column width="50%"}
What is the **A**verage **Treatment** **E**ffect of Smoking 🚬 (vs not smoking 🚭) for Lung Cancer 🦀?

Need for confounder adjustment:

1.  Parent that smokes 😤 (unlike non-smoker parent 😐) increases the probability of a child that smokes 🚬.

2.  Parent that smokes 😤 increases the probability of a lung cancer for a child through passive smoking.
:::

::: {.column width="50%"}
![](dag_lecture.svg)
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}
countefactual_data <- tibble::tribble(
   ~"X", ~"A", ~"Y_a0", ~"Y_a1",
   "0", "1", "0", "1",
   "0", "1", "0", "0",
   "0", "0", "1", "1",
   "0", "0", "0", "1",
   "0", "0", "0", "0",
   "0", "0", "0", "0",
   "1", "1", "1", "1",
   "1", "1", "1", "1",
   "1", "1", "1", "1",
   "1", "1", "1", "1",
   "1", "1", "1", "1",
   "1", "1", "0", "1",
   "1", "1", "0", "0",
   "1", "1", "0", "0",
   "1", "0", "1", "1",
   "1", "0", "0", "1"
  ) |> 
  mutate(
    observed_y = case_when(
      A == "0" ~ Y_a0,
      A == "1" ~ Y_a1
    ),
    ITE = as.numeric(Y_a1) - as.numeric(Y_a0)
  ) |> 
  arrange(
    A, observed_y, X
  )

  

countefactual_data_reactable_raw <- countefactual_data |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  columns = list(
    observed_y = colDef(
      show = FALSE
    ),
    ITE = colDef(
      show = FALSE
    ),
    Y_a0 = colDef(
      name = "Y<sup>0",
      html = TRUE
    ),
    Y_a1 = colDef(
      name = "Y<sup>1",
      html = TRUE
    )
  ),
  defaultColDef = colDef(maxWidth = 40),
  style = list(fontSize = "1rem"))


countefactual_data_reactable_raw
```
:::

::: {.column width="65%"}
#### 😇 God-Given Counterfactual Data

-   By God-Given knowledge we have access to the true counterfactual outcomes for each individual for smoking 🚬 and for not smoking 🚭
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


countefactual_data_reactable_raw_with_ITE <- countefactual_data |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  columns = list(
    observed_y = colDef(
      show = FALSE
    ),
    Y_a0 = colDef(
      name = "Y<sup>0",
      html = TRUE
    ),
    Y_a1 = colDef(
      name = "Y<sup>1",
      html = TRUE
    ),
    ITE = colDef(
      style = function(value) {
      if(value == "1")
      list(background = "pink")
    }
    )
  ),
  defaultColDef = colDef(maxWidth = 40),
  style = list(fontSize = "1rem"))

countefactual_data_reactable_raw_with_ITE
```
:::

::: {.column width="65%"}
#### 😇 God-Given Counterfactual Data

-   By God-Given knowledge we have access to the true counterfactual outcomes for each individual for smoking 🚬 and for not smoking 🚭

-   We can also calculate the Causal Individual Treatment Effect and the true Causal Average Treatment Effect:

$\text{Causal ATE} = \frac{11}{16} - \frac{7}{16} = \frac{4}{16} = 0.25$
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


countefactual_data_reactable_raw_with_observed_outcome <- countefactual_data |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  columns = list(
    observed_y = colDef(
      name = "Y"
    ),
    Y_a0 = colDef(
      name = "Y<sup>0",
      html = TRUE
    ),
    Y_a1 = colDef(
      name = "Y<sup>1",
      html = TRUE
    ),
    ITE = colDef(
      show = FALSE
    )
  ),
  defaultColDef = colDef(maxWidth = 40),
  style = list(fontSize = "1rem"))

countefactual_data_reactable_raw_with_observed_outcome
```
:::

::: {.column width="65%"}
#### 😇 God-Given Counterfactual Data

-   By God-Given knowledge we have access to the true counterfactual outcomes for each individual for smoking 🚬 and for not smoking 🚭

-   We can also calculate the Causal Individual Treatment Effect and the true Causal Average Treatment Effect:

$\text{Causal ATE} = \frac{11}{16} - \frac{7}{16} = \frac{4}{16} = 0.25$

-   The observed outcome is the suitable counterfactual outcome to the given treatment.
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_reactable_raw <- countefactual_data |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  columns = list(
    observed_y = colDef(
      name = "Y",
      style = function(value) {
      if(value == "1")
      list(background = "pink")
    }
    ),
    A = colDef(
      style = function(value) {
      if (value == "1")
      list(background = "#F2F3AE")
      } 
    ),
    Y_a0 = colDef(
      show = FALSE
    ),
    Y_a1 = colDef(
      show = FALSE
    ),
    ITE = colDef(
      show = FALSE
    )
  ),
  defaultColDef = colDef(maxWidth = 40),
  style = list(fontSize = "1rem"))

real_data_reactable_raw
```
:::

::: {.column width="65%"}
#### 😈 Observed Data

-   By God-Given knowledge we have access to the true counterfactual outcomes for each individual for smoking 🚬 and for not smoking 🚭

-   We can also calculate the Causal Individual Treatment Effect and the true Causal Average Treatment Effect:

$\text{Causal ATE} = \frac{11}{16} - \frac{7}{16} = \frac{4}{16} = 0.25$

-   The observed outcome is the suitable counterfactual outcome to the given treatment.

-   In real life we can analyse only the observed outcomes.

The naive Associational Average Treatment Effect will be:

$\text{Associational ATE} = \frac{7}{10} - \frac{2}{6} \approx 0.366$

Which is obviously biased.
:::
:::

# First Solution: IPW ⚖️

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_no_outcome <- countefactual_data |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  columns = list(
    observed_y = colDef(
      name = "Y",
      show = FALSE
    ),
    Y_a0 = colDef(
      show = FALSE
    ),
    Y_a1 = colDef(
      show = FALSE
    ),
    ITE = colDef(
      show = FALSE
    )
  ),
  defaultColDef = colDef(maxWidth = 40),
  style = list(fontSize = "1rem"))

real_data_no_outcome
```
:::

::: {.column width="65%"}
#### IPW ⚖️

-   IPW is a method for manipulating the sample while creating pseudo-population. Therefore, we don't need the outcome for the first step.
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_no_outcome <- countefactual_data |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  columns = list(
    observed_y = colDef(
      name = "Y",
      show = FALSE
    ),
    Y_a0 = colDef(
      show = FALSE
    ),
    Y_a1 = colDef(
      show = FALSE
    ),
    ITE = colDef(
      show = FALSE
    )
  ),
  defaultColDef = colDef(maxWidth = 40),
  style = list(fontSize = "1rem"))

real_data_no_outcome
```
:::

::: {.column width="65%"}
#### IPW ⚖️

-   IPW is a method for manipulating the sample while creating pseudo-population. Therefore, we don't need the outcome for the first step.

-   But we do need to estimate the probability of having a treatment for each individual given the confounders.

$p(A|X)$

On the non-parametric case it's just the proportions of the treated patients from the subpopulation $X=x$
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_no_outcome_a0_x0 <- countefactual_data |> 
  select(X, A) |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  defaultColDef = colDef(maxWidth = 40),
  rowStyle = function(index) {
    if (
      (countefactual_data[index, "X"] == "0") &&
      (countefactual_data[index, "A"] == "0")) list(background = "rgba(0, 0, 0, 0.15)") else if (
        (countefactual_data[index, "X"] == "0") &&
      (countefactual_data[index, "A"] == "1")
      ) list(background = "rgba(0, 0, 0, 0.05)")
  },
  style = list(fontSize = "1rem"), 
  fullWidth = FALSE)

real_data_no_outcome_a0_x0
```
:::

::: {.column width="65%"}
#### IPW ⚖️

-   IPW is a method for manipulating the sample while creating pseudo-population. Therefore, we don't need the outcome for the first step.

-   But we do need to estimate the probability of having a treatment for each individual given the confounders.

$p(A = 0|X = 0) = \frac{4}{6}$
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_no_outcome_a0_x0 <- countefactual_data |> 
  select(X, A) |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  defaultColDef = colDef(maxWidth = 40),
  rowStyle = function(index) {
    if (
      (countefactual_data[index, "X"] == "0") &&
      (countefactual_data[index, "A"] == "1")) list(background = "rgba(0, 0, 0, 0.15)") else if (
        (countefactual_data[index, "X"] == "0") &&
      (countefactual_data[index, "A"] == "0")
      ) list(background = "rgba(0, 0, 0, 0.05)")
  },
  style = list(fontSize = "1rem"), 
  fullWidth = FALSE)

real_data_no_outcome_a0_x0
```
:::

::: {.column width="65%"}
#### IPW ⚖️

-   IPW is a method for manipulating the sample while creating pseudo-population. Therefore, we don't need the outcome for the first step.

-   But we do need to estimate the probability of having a treatment for each individual given the confounders.

$p(A = 0|X = 0) = \frac{4}{6}$

$p(A = 1|X = 0) = \frac{2}{6}$
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_no_outcome_a0_x0 <- countefactual_data |> 
  select(X, A) |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  defaultColDef = colDef(maxWidth = 40),
  rowStyle = function(index) {
    if (
      (countefactual_data[index, "X"] == "1") &&
      (countefactual_data[index, "A"] == "0")) list(background = "rgba(0, 0, 0, 0.15)") else if (
        (countefactual_data[index, "X"] == "1") &&
      (countefactual_data[index, "A"] == "1")
      ) list(background = "rgba(0, 0, 0, 0.05)")
  },
  style = list(fontSize = "1rem"), 
  fullWidth = FALSE)

real_data_no_outcome_a0_x0
```
:::

::: {.column width="65%"}
#### IPW ⚖️

-   IPW is a method for manipulating the sample while creating pseudo-population. Therefore, we don't need the outcome for the first step.

-   But we do need to estimate the probability of having a treatment for each individual given the confounders.

$p(A = 0|X = 0) = \frac{4}{6}$

$p(A = 1|X = 0) = \frac{2}{6}$

$p(A = 0|X = 1) = \frac{2}{10}$
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_no_outcome_a0_x0 <- countefactual_data |> 
  select(X, A) |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  defaultColDef = colDef(maxWidth = 40),
  rowStyle = function(index) {
    if (
      (countefactual_data[index, "X"] == "1") &&
      (countefactual_data[index, "A"] == "1")) list(background = "rgba(0, 0, 0, 0.15)") else if (
        (countefactual_data[index, "X"] == "1") &&
      (countefactual_data[index, "A"] == "0")
      ) list(background = "rgba(0, 0, 0, 0.05)")
  },
  style = list(fontSize = "1rem"), 
  fullWidth = FALSE)

real_data_no_outcome_a0_x0
```
:::

::: {.column width="65%"}
#### IPW ⚖️

-   IPW is a method for manipulating the sample while creating pseudo-population. Therefore, we don't need the outcome for the first step.

-   But we do need to estimate the probability of having a treatment for each individual given the confounders.

$p(A = 0|X = 0) = \frac{4}{6}$

$p(A = 1|X = 0) = \frac{2}{6}$

$p(A = 0|X = 1) = \frac{2}{10}$

$p(A = 1|X = 1) = \frac{8}{10}$
:::
:::

## Weights for Non-Smoking Parents 😐 {.smaller}

::: columns
::: {.column width="50%"}
```{r}

data.frame(
  Population = c(
    paste(
    rep(
    "🚭", 4
  ), collapse = ""),
  paste(
    rep(
    "🚬", 2
  ), collapse = "")
    ),
  Propensity = c(
    "4 / 6",
    "2 / 6"
  ),
  IPW = c(
    "6 / 4",
    "6 / 2"
  )
  ) |>
  reactable(
    columns = list(
      Population = colDef(
        maxWidth = 160
      ),
      Propensity = colDef(
        name = "P(A|L)",
        maxWidth = 100
      ),
      IPW = colDef(
        name = "W<sup>A",
        maxWidth = 100,
        html = TRUE
      )
      ),
  columnGroups = list(
    colGroup(name = "Original Sample", columns = c("Population", "Propensity"))
  )
  )

```
:::

::: {.column width="50%"}
![](imbalanced_non_smoking_parents.svg)
:::
:::

## Weights for Non-Smoking Parents 😐 {.smaller}

::: columns
::: {.column width="50%"}
```{r}

data.frame(
  Population = c(
    paste(
    rep(
    "🚭", 6
  ), collapse = ""),
  paste(
    rep(
    "🚬", 6
  ), collapse = "")
    ),
  Propensity = c(
    "6 / 12",
    "6 / 12"
  )
  ) |>
  reactable(
    columns = list(
      Population = colDef(
        maxWidth = 160
      ),
      Propensity = colDef(
        name = "P(A|L)",
        maxWidth = 100
      )
      ),
  columnGroups = list(
    colGroup(name = "Weighted Sample", columns = c("Population", "Propensity"))
  )
  )

```
:::

::: {.column width="50%"}
![](balanced_non_smoking_parents.svg)
:::
:::

## Weights for Smoking Parents 😤 {.smaller}

::: columns
::: {.column width="50%"}
```{r}

data.frame(
  Population = c(
    paste(
    rep(
    "🚭", 2
  ), collapse = ""),
  paste(
    rep(
    "🚬", 8
  ), collapse = "")
    ),
  Propensity = c(
    "2 / 10",
    "8 / 10"
  ),
  IPW = c(
    "10 / 2",
    "10 / 8"
  )
  ) |>
  reactable(
    columns = list(
      Population = colDef(
        maxWidth = 160
      ),
      Propensity = colDef(
        name = "P(A|L)",
        maxWidth = 100
      ),
      IPW = colDef(
        name = "W<sup>A",
        maxWidth = 100,
        html = TRUE
      )
      ),
  columnGroups = list(
    colGroup(name = "Original Sample", columns = c("Population", "Propensity"))
  )
  )

```
:::

::: {.column width="50%"}
![](imbalanced_smoking_parents.svg)
:::
:::

## Weights for Smoking Parents 😤 {.smaller}

::: columns
::: {.column width="50%"}
```{r}

data.frame(
  Population = c(
    paste(
    rep(
    "🚭", 10
  ), collapse = ""),
  paste(
    rep(
    "🚬", 10
  ), collapse = "")
    ),
  Propensity = c(
    "10 / 20",
    "10 / 20"
  )
  ) |>
  reactable(
    columns = list(
      Population = colDef(
        maxWidth = 160
      ),
      Propensity = colDef(
        name = "P(A|L)",
        maxWidth = 100
      )
      ),
  columnGroups = list(
    colGroup(name = "Weighted Sample", columns = c("Population", "Propensity"))
  )
  )

```
:::

::: {.column width="50%"}
![](balanced_smoking_parents.svg)
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_with_weights <- countefactual_data |> 
  select(X, A, observed_y) |>
  mutate(W_a = 
           case_when(
             (X=="0" & A == "0") ~ 1.5,
             (X=="0" & A == "1") ~ 3,
             (X=="1" & A == "0") ~ 5,
             (X=="1" & A == "1") ~1.25
           )) |> 
  reactable(
  columns = list(
    W_a = colDef(
      name = "W<sup>a",
      html = TRUE
    ),
    observed_y = colDef(
      name = "Y"
    )
  ),
  rowStyle = function(index) {
    if (
      (countefactual_data[index, "observed_y"] == "1") &&
      (countefactual_data[index, "A"] == "0")) list(background = "rgba(0, 0, 0, 0.15)") else if (
        (countefactual_data[index, "observed_y"] == "0") &&
      (countefactual_data[index, "A"] == "0")
      ) list(background = "rgba(0, 0, 0, 0.05)")
  },
  sortable = FALSE,
  defaultPageSize = 19,
  defaultColDef = colDef(maxWidth = 45),
  style = list(fontSize = "1rem"), 
  fullWidth = FALSE) 
  

real_data_with_weights
```
:::

::: {.column width="65%"}
#### Calculate estimate for Causal ATE ⚖️

##### Non-Smokers outcome proportion (Psuedo-population)

-   6.5 cases (half a person is not a problem in pseudo population) of lung out of 16.
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_with_weights <- countefactual_data |> 
  select(X, A, observed_y) |>
  mutate(W_a = 
           case_when(
             (X=="0" & A == "0") ~ 1.5,
             (X=="0" & A == "1") ~ 3,
             (X=="1" & A == "0") ~ 5,
             (X=="1" & A == "1") ~1.25
           )) |> 
  reactable(
  columns = list(
    W_a = colDef(
      name = "W<sup>a",
      html = TRUE
    ),
    observed_y = colDef(
      name = "Y"
    )
  ),
  rowStyle = function(index) {
    if (
      (countefactual_data[index, "observed_y"] == "1") &&
      (countefactual_data[index, "A"] == "1")) list(background = "rgba(0, 0, 0, 0.15)") else if (
        (countefactual_data[index, "observed_y"] == "0") &&
      (countefactual_data[index, "A"] == "1")
      ) list(background = "rgba(0, 0, 0, 0.05)")
  },
  sortable = FALSE,
  defaultPageSize = 19,
  defaultColDef = colDef(maxWidth = 45),
  style = list(fontSize = "1rem"), 
  fullWidth = FALSE) 
  

real_data_with_weights
```
:::

::: {.column width="65%"}
#### Calculate estimate for Causal ATE ⚖️

##### Non-Smokers outcome proportion (Psuedo-population)

-   6.5 cases (half a person is not a problem in pseudo population) of lung out of 16.

##### Smokers Outcome Proporiton (Psuedo-population)

-   10.5 cases out of 16.
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_with_weights <- countefactual_data |> 
  select(X, A, observed_y) |>
  mutate(W_a = 
           case_when(
             (X=="0" & A == "0") ~ 1.5,
             (X=="0" & A == "1") ~ 3,
             (X=="1" & A == "0") ~ 5,
             (X=="1" & A == "1") ~1.25
           )) |> 
  reactable(
  columns = list(
    W_a = colDef(
      name = "W<sup>a",
      html = TRUE
    ),
    observed_y = colDef(
      name = "Y"
    )
  ),
  sortable = FALSE,
  defaultPageSize = 19,
  defaultColDef = colDef(maxWidth = 45),
  style = list(fontSize = "1rem"), 
  fullWidth = FALSE) 
  

real_data_with_weights
```
:::

::: {.column width="65%"}
#### Calculate estimate for Causal ATE ⚖️

##### Non-Smokers outcome proportion (Psuedo-population)

-   6.5 cases (half a person is not a problem in pseudo population) of lung out of 16.

##### Smokers Outcome Proporiton (Psuedo-population)

-   10.5 cases out of 16.

##### Estimate of Causal Average Treatment Effect

$\hat{ATE} = \frac{10.5}{16} - \frac{6.5}{16} = 0.25$
:::
:::

# Second Solution: Standardization 🗃

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


real_data_reactable_raw
```
:::

::: {.column width="65%"}
#### Standardization 🗃

-   The marginal counterfactual risk $Pr[Y^{a=1}]$ is the weighted average of the stratum-specific risks $Pr[Y^{a=1}|X =0]$ and $Pr[Y^{a=1}|X =1]$, in other words:

$Pr[Y^{a=1}] = \sum_{X=x}Pr[Y^{a=1}|X =x]*Pr(X=x)$

-   Under conditional exchangeability:

$\hat{Pr[Y^{a=1}]} = \sum_{X=x}Pr[Y=1|X =x, A=1]*Pr(X=x)$

-   Therefore we can estimate directly the estimate for the average treatment effect if we will use estimates for every stratum of the confounders (X) and the Treatment (A).

-   The same goes for:

$\hat{Pr[Y^{a=0}]} = \sum_{X=x}Pr[Y=1|X =x, A=0]*Pr(X=x)$
:::
:::

##  {.smaller}

::: columns
::: {.column width="35%"}
```{r}


countefactual_data |> 
  arrange(X) |> 
  reactable(
  sortable = FALSE,
  defaultPageSize = 19,
  columns = list(
    observed_y = colDef(
      name = "Y",
      style = function(value) {
      if(value == "1")
      list(background = "pink")
    }
    ),
    A = colDef(
      style = function(value) {
      if (value == "1")
      list(background = "#F2F3AE")
      } 
    ),
    X = colDef(
      style = function(value) {
      if (value == "1")
      list(background = "lightblue")
      } 
    ),
    Y_a0 = colDef(
      show = FALSE
    ),
    Y_a1 = colDef(
      show = FALSE
    ),
    ITE = colDef(
      show = FALSE
    )
  ),
  defaultColDef = colDef(maxWidth = 40),
  style = list(fontSize = "1rem"))

```
:::

::: {.column width="65%"}
#### Standardization 🗃

$\hat{Pr[Y^{a=1}]} = \frac{1}{2}*\frac{6}{16} + \frac{6}{8}*\frac{10}{16} = 0.625$

$\hat{Pr[Y^{a=0}]} = \frac{1}{4}*\frac{6}{16} + \frac{1}{2}*\frac{10}{16} = 0.425$

##### Estimate of Causal Average Treatment Effect

$\hat{ATE} = \frac{10.5}{16} - \frac{6.5}{16} = 0.25$
:::
:::

## Parametric Versions

-   Parametric versions of IPW and Standardization will yield different estimates unlike the Non-Parametric Versions.

-   For IPW: We can use ML for propensity! 🤖

-   For Standardization: We can use ML 🤖 (T/S/X Learners) or old-school outcome regressions 👴 (just don't use colliders).

## Benefits

-   **IPW**: Model for the treatment without overfitting the for the outcome.

-   **Standardization**: CATE / ITE.

-   **Doubly Robust**: Estimates are consistent even if we are wrong about treatment model or outcome model.

[Check out our paper!](https://pubmed.ncbi.nlm.nih.gov/35598055/)
